When Imitation Learning Outperforms Reinforcement Learning in Surgical Action Planning

### 2025-10-23 docs/paper_manuscript/results/2025-07-05_21-17-49/fold0/miccai_paper/archives/paper.tex
clean code

\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Teleoperated robotic surgery provides a natural interface to acquire expert demonstrations for imitation learning (IL), while reinforcement learning (RL) could in principle discover new strategies and achieve beyond expert-level performance. However, the question of when to use IL versus RL in surgical domains remains largely unexplored. We conducted a comprehensive comparison of IL versus RL approaches for surgical action planning on the CholecT50 dataset. Our baseline uses supervised learning on expert demonstration videos to learn surgical behavior through direct observation. We systematically evaluated: (1) standard IL with causal prediction, (2) RL with learned rewards via inverse RL, and (3) world model-based RL with forward simulation. Our IL baseline achieves 34.6\% current action mAP and 33.6\% next action mAP with smooth planning degradation (33.6\% at 1s to 29.2\% at 10s). Surprisingly, sophisticated RL approaches failed to improve upon this baseline. We found that distribution matching on the evaluation test set favors the IL baseline over potentially valid or even better new policies that differ from expert demonstrations used for training. This challenges assumptions about method hierarchy and provides crucial insights for surgical AI development.

@@

\textbf{Contributions}: (1) First systematic comparison of IL vs RL for surgical action planning, (2) Important negative result showing when RL doesn't help in expert domains, (3) Comprehensive evaluation framework for temporal surgical planning, and (4) Analysis of how expert data quality and evaluation metrics affect method selection in surgical domains.

% \begin{abstract}
% Surgical action triplet prediction has primarily focused on recognition tasks for activity analysis. However, real-time surgical assistance requires next action prediction for planning and control applications. This work presents the first comprehensive comparison of Imitation Learning (IL) versus Reinforcement Learning (RL) approaches for surgical next action prediction, evaluating both recognition accuracy and planning capability. We propose an autoregressive IL baseline achieving 0.979 mAP on CholecT50, and systematically compare it against multiple RL variants including world model-based RL, direct video RL, and inverse RL enhancement. Our analysis reveals that while IL excels at single-step prediction, RL provides scenario-specific improvements for complex surgical situations requiring multi-step reasoning. These findings inform optimal AI approach selection for clinical surgical assistance applications.

% \keywords{Surgical AI \and Action Prediction \and Imitation Learning \and Reinforcement Learning \and Planning}
% \end{abstract}

@@

\begin{table}[h]
\centering
\caption{Comparative Results: IL vs RL Approaches for Surgical Action Planning. All values are IVT mAP scores in \%. mAP$_{t=0}$ refers to current action recognition, mAP$_{t+1s}$ to 1-second prediction, etc.}
\label{tab:main_results}
\begin{tabular}{lcccc}
\toprule
\textbf{Method} & \textbf{mAP$_{t=0}$} & \textbf{mAP$_{t+1s}$} & \textbf{mAP$_{t+5s}$} & \textbf{mAP$_{t+10s}$} \\
\midrule
DARIL (Ours) & \textbf{34.6} & \textbf{33.6} & \textbf{31.2} & \textbf{29.2} \\
\midrule
DARIL + Direct Video RL & \textit{33.2} & \textit{22.6} & \textit{19.3} & \textit{15.9} \\
DARIL + IRL & \textit{33.1} & \textit{32.1} & \textit{29.6} & \textit{28.1} \\
Latent World Model + RL & \textit{33.1} & \textit{14.0} & \textit{9.1} & \textit{3.1} \\
\bottomrule
\end{tabular}
\end{table}

>>>>

\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Surgical action triplet prediction has primarily focused on recognition tasks for activity analysis. However, real-time surgical assistance requires next action prediction for planning and control applications. Teleoperated robotic surgery provides a natural interface to acquire expert demonstrations for imitation learning (IL), while reinforcement learning (RL) could in principle discover new strategies and achieve beyond expert-level performance. The question of when to use IL versus RL in surgical domains remains largely unexplored. We conducted a comprehensive comparison of IL versus RL approaches for surgical action planning on the CholecT50 dataset, evaluating both recognition accuracy and planning capability. Our baseline uses supervised learning on expert demonstration videos to learn surgical behavior through direct observation. We systematically evaluated: (1) standard IL with causal prediction, (2) RL with learned rewards via inverse RL, and (3) world model-based RL with forward simulation. Our IL baseline achieves 34.6\% current action mAP and 33.6\% next action mAP with smooth planning degradation (33.6\% at 1s to 29.2\% at 10s). Surprisingly, sophisticated RL approaches failed to improve upon this baseline. We found that distribution matching on the evaluation test set favors the IL baseline over potentially valid or even better new policies that differ from expert demonstrations used for training. This challenges assumptions about method hierarchy and provides crucial insights for surgical AI development.

@@

\textbf{Contributions}: (1) \textit{First systematic IL vs RL comparison for surgical action planning}: We provide the first comprehensive evaluation comparing imitation learning and reinforcement learning approaches for surgical action planning, addressing a fundamental methodological question in surgical AI. (2) \textit{Surprising negative results with clear performance gaps}: We demonstrate that sophisticated RL methods consistently underperform our IL baseline—world model RL drops to 3.1\% mAP at 10s vs 29.2\% for DARIL, while direct video RL achieves only 15.9\% vs 29.2\%, challenging assumptions about RL superiority in sequential decision making. (3) \textit{Novel DARIL architecture achieving strong temporal consistency}: Our dual-task autoregressive imitation learning approach maintains robust performance across planning horizons (34.6\% current action mAP degrading smoothly to 29.2\% at 10-second planning), establishing a new baseline for surgical action planning. (4) \textit{Insights into evaluation bias and domain characteristics}: We identify how expert demonstration alignment with test distributions systematically favors IL over potentially valid RL policies, providing crucial guidance for evaluation design and method selection in expert domains like surgery.

@@

\begin{table}[h]
\centering
\caption{Comparative Results: IL vs RL Approaches for Surgical Action Planning. All values are IVT mAP scores in \%. mAP$_{t=0}$ refers to current action recognition, mAP$_{t+1s}$ to 1-second prediction, etc.}
\label{tab:main_results}
\begin{tabular}{lcccc}
\toprule
\textbf{Method} & \textbf{mAP$_{t=0}$} & \textbf{mAP$_{t+1s}$} & \textbf{mAP$_{t+5s}$} & \textbf{mAP$_{t+10s}$} \\
\midrule
DARIL (Ours) & \textbf{34.6} & \textbf{33.6} & \textbf{31.2} & \textbf{29.2} \\
\midrule
DARIL + Direct Video RL & 33.2 & 22.6 & 19.3 & 15.9 \\
DARIL + IRL & 33.1 & 32.1 & 29.6 & 28.1 \\
Latent World Model + RL & 33.1 & 14.0 & 9.1 & 3.1 \\
\bottomrule
\end{tabular}
\end{table}

### 2025-10-23 docs/paper_manuscript/results/2025-07-05_21-17-49/fold0/miccai_paper/paper_manuscript copy.tex
clean code

\maketitle
%
\begin{abstract}
Surgical action planning requires predicting future instrument-verb-target triplets for real-time assistance. While teleoperated robotic surgery provides natural expert demonstrations for imitation learning (IL), reinforcement learning (RL) could potentially discover superior strategies through exploration. We present the first comprehensive comparison of IL versus RL for surgical action planning on CholecT50. Our Dual-task Autoregressive Imitation Learning (DARIL) baseline achieves 34.6\% action triplet recognition mAP and 33.6\% next frame prediction mAP with smooth planning degradation to 29.2\% at 10-second horizons. We evaluated three RL variants: world model-based RL, direct video RL, and inverse RL enhancement. Surprisingly, all RL approaches underperformed DARIL—world model RL dropped to 3.1\% mAP at 10s while direct video RL achieved only 15.9\%. Our analysis reveals that distribution matching on expert-annotated test sets systematically favors IL over potentially valid RL policies that differ from training demonstrations. This challenges assumptions about RL superiority in sequential decision making and provides crucial insights for surgical AI development.

@@

Surgical action planning, predicting future instrument-verb-target relationships in surgical videos, represents a critical component for real-time surgical assistance systems. While prior work has predominantly focused on recognition tasks~\cite{nwoye2022cholect50,nwoye2020recognition,nwoye2023cholectriplet2021}, prospective action planning presents unique challenges requiring multi-horizon prediction capabilities under safety-critical constraints.

@@

\textbf{Contributions}: (1) \textit{First systematic IL vs RL comparison}: Comprehensive evaluation addressing fundamental methodological questions in surgical AI. (2) \textit{Surprising negative results}: RL methods consistently underperform IL with world model RL dropping to 3.1\% mAP vs 29.2\% for DARIL at 10s planning. (3) \textit{Novel DARIL architecture}: Dual-task autoregressive approach maintaining robust temporal consistency (34.6\% action triplet recognition degrading smoothly to 29.2\% at 10s). (4) \textit{Evaluation bias insights}: Analysis of how expert demonstration alignment systematically favors IL over valid RL policies.

@@

Given surgical video frames $\{f_1, f_2, ..., f_t\}$, we predict future action triplets $\{a_{t+1}, a_{t+2}, ..., a_{t+H}\}$ where $H$ represents the planning horizon. Each triplet $a_i = (I_i, V_i, T_i)$ consists of instrument, verb, and target components from predefined vocabularies. Importantly, each frame can contain multiple simultaneous actions (0-3 per frame) due to multiple robotic arms with different instruments, making the problem highly sparse with 100 distinct action classes total. We evaluate both single-step prediction ($H=1$) for recognition and multi-step prediction ($H>1$) for planning assessment.

@@

\begin{equation}
p(a_{t+1}|f_{1:t}) = \text{GPT-2}(\text{FrameEmb}(f_{1:t}))
\end{equation}

\textbf{Architecture}: The model processes 1024-dimensional Swin transformer features~\cite{liu2021swin} through: (1) BiLSTM encoder for temporal current action recognition, (2) GPT-2 decoder~\cite{radford2019language} for causal future action generation using only frame embeddings as input, and (3) separate prediction heads for instrument, verb, and target components.

@@

Table~\ref{tab:main_results} presents our experimental findings. DARIL achieves 34.6\% action triplet recognition mAP and 33.6\% next frame prediction mAP, consistently outperforming all RL variants across planning horizons. The performance gaps are substantial—world model RL drops to 3.1\% at 10s while DARIL maintains 29.2\%.

\begin{table}[h]
\centering
\caption{IL vs RL Performance Comparison. All values are IVT mAP (\%). Current refers to action triplet recognition, 1s/5s/10s refer to next frame prediction at different horizons. Statistical significance (p<0.05) confirmed via paired t-tests across video splits.}
\label{tab:main_results}
\begin{tabular}{lcccc}
\toprule
\textbf{Method} & \textbf{Current} & \textbf{1s} & \textbf{5s} & \textbf{10s} \\
\midrule
DARIL (Ours) & \textbf{34.6} & \textbf{33.6} & \textbf{31.2} & \textbf{29.2} \\
DARIL + IRL & 33.1 & 32.1 & 29.6 & 28.1 \\
DARIL + Direct Video RL & 33.2 & 22.6 & 19.3 & 15.9 \\
Latent World Model + RL & 33.1 & 14.0 & 9.1 & 3.1 \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{table}[h]
\centering
\caption{DARIL Component-wise Performance Analysis. Current refers to action triplet recognition, Next refers to next frame prediction.}
\label{tab:component_analysis}
\begin{tabular}{lcccccc}
\toprule
\textbf{Component} & \textbf{Recognition} & \textbf{Next Frame} & \textbf{Component} & \textbf{Recognition} & \textbf{Next Frame} \\
\midrule
Instrument (I) & 91.4 & 88.2 & Instrument-Verb (IV) & 42.9 & 38.8 \\
Verb (V) & 69.4 & 68.1 & Instrument-Target (IT) & 43.5 & 43.6 \\
Target (T) & 52.7 & 52.5 & IVT & 34.6 & 33.6 \\
\bottomrule
\end{tabular}
\end{table}

@@

Figure~\ref{fig:planning_analysis} demonstrates DARIL's smooth degradation across horizons from 33.6\% at 1s to 29.2\% at 10s (13.1\% relative decrease). This contrasts with RL approaches showing steeper degradation patterns, particularly world model RL exhibiting catastrophic performance loss.

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{planning_analysis_simple.png}
\caption{Planning performance comparison across methods and time horizons. DARIL maintains superior and stable performance while RL approaches show varying degrees of degradation. Error bars indicate 95\% confidence intervals.}
\label{fig:planning_analysis}
\end{figure}

@@

\textbf{Bootstrapping RL with IL}: Our findings suggest a promising approach for surgical AI: bootstrapping RL models with IL-learned basic skills, then using physics simulators or world models for safe exploration of new techniques. This addresses the fundamental challenge that exploration and mistakes are crucial for improving surgical techniques, but must be tested in simulation rather than on patients.

@@

Several limitations should be considered: (1) Single dataset evaluation on CholecT50 may not generalize to other surgical procedures. (2) Expert test data similar to training distributions may favor IL—results might differ with sub-expert or out-of-distribution scenarios. (3) Evaluation metrics directly reward expert-like behavior—alternative criteria focusing on patient outcomes might favor RL. (4) More sophisticated RL implementations with better state representations and reward design might outperform IL. (5) \textbf{Overfitting concerns}: Our models likely overfit to the limited number of videos and may not generalize well to test videos, indicating need for larger datasets and better simulators, whether world models or physics engines. (6) \textbf{Lack of reward feedback}: Our RL approaches suffered from insufficient reward signals from possible future states and lack of outcome data, limiting their learning effectiveness.

@@

%
\bibliographystyle{splncs04}

>>>>

\maketitle
%
\begin{abstract}
Surgical action planning requires predicting future instrument-verb-target triplets for real-time assistance. While teleoperated robotic surgery provides natural expert demonstrations for imitation learning (IL), reinforcement learning (RL) could potentially discover superior strategies. We present the first comprehensive comparison of IL versus RL for surgical action planning on CholecT50. Our Dual-task Autoregressive Imitation Learning (DARIL) baseline achieves 34.6\% current action mAP and 33.6\% next action mAP with smooth planning degradation to 29.2\% at 10-second horizons. We evaluated three RL variants: world model-based RL, direct video RL, and inverse RL enhancement. Surprisingly, all RL approaches underperformed DARIL—world model RL dropped to 3.1\% mAP at 10s while direct video RL achieved only 15.9\%. Our analysis reveals that distribution matching on expert-annotated test sets systematically favors IL over potentially valid RL policies that differ from training demonstrations. This challenges assumptions about RL superiority in sequential decision making and provides crucial insights for surgical AI development.

@@

Surgical action planning—predicting future instrument-verb-target relationships in surgical videos—represents a critical component for real-time surgical assistance systems. While prior work has predominantly focused on recognition tasks~\cite{nwoye2022cholect50,nwoye2020recognition,nwoye2023cholectriplet2021}, prospective action planning presents unique challenges requiring multi-horizon prediction capabilities under safety-critical constraints.

@@

\textbf{Contributions}: (1) \textit{First systematic IL vs RL comparison}: Comprehensive evaluation addressing fundamental methodological questions in surgical AI. (2) \textit{Surprising negative results}: RL methods consistently underperform IL—world model RL drops to 3.1\% mAP vs 29.2\% for DARIL at 10s planning. (3) \textit{Novel DARIL architecture}: Dual-task autoregressive approach maintaining robust temporal consistency (34.6\% to 29.2\% across horizons). (4) \textit{Evaluation bias insights}: Analysis of how expert demonstration alignment systematically favors IL over valid RL policies.

@@

Given surgical video frames $\{f_1, f_2, ..., f_t\}$, we predict future action triplets $\{a_{t+1}, a_{t+2}, ..., a_{t+H}\}$ where $H$ represents the planning horizon. Each triplet $a_i = (I_i, V_i, T_i)$ consists of instrument, verb, and target components from predefined vocabularies. We evaluate both single-step prediction ($H=1$) for recognition and multi-step prediction ($H>1$) for planning assessment.

@@

\begin{equation}
p(a_{t+1}|f_{1:t}, a_{1:t}) = \text{GPT-2}(\text{FrameEmb}(f_{1:t}))
\end{equation}

\textbf{Architecture}: The model processes 1024-dimensional Swin transformer features~\cite{liu2021swin} through: (1) BiLSTM encoder for temporal current action recognition, (2) GPT-2 decoder~\cite{radford2019language} for causal future action generation, and (3) separate prediction heads for instrument, verb, and target components.

@@

Table~\ref{tab:main_results} presents our experimental findings. DARIL achieves 34.6\% current action mAP and 33.6\% next action mAP, consistently outperforming all RL variants across planning horizons. The performance gaps are substantial—world model RL drops to 3.1\% at 10s while DARIL maintains 29.2\%.

\begin{table}[h]
\centering
\caption{IL vs RL Performance Comparison. All values are IVT mAP (\%). Statistical significance (p<0.05) confirmed via paired t-tests across video splits.}
\label{tab:main_results}
\begin{tabular}{lcccc}
\toprule
\textbf{Method} & \textbf{Current} & \textbf{1s} & \textbf{5s} & \textbf{10s} \\
\midrule
DARIL (Ours) & \textbf{34.6} & \textbf{33.6} & \textbf{31.2} & \textbf{29.2} \\
DARIL + IRL & 33.1 & 32.1 & 29.6 & 28.1 \\
DARIL + Direct Video RL & 33.2 & 22.6 & 19.3 & 15.9 \\
Latent World Model + RL & 33.1 & 14.0 & 9.1 & 3.1 \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{table}[h]
\centering
\caption{DARIL Component-wise Performance Analysis}
\label{tab:component_analysis}
\begin{tabular}{lcccccc}
\toprule
\textbf{Component} & \textbf{Current} & \textbf{Next} & \textbf{Component} & \textbf{Current} & \textbf{Next} \\
\midrule
Instrument (I) & 91.4 & 88.2 & Instrument-Verb (IV) & 42.9 & 38.8 \\
Verb (V) & 69.4 & 68.1 & Instrument-Target (IT) & 43.5 & 43.6 \\
Target (T) & 52.7 & 52.5 & IVT & 34.6 & 33.6 \\
\bottomrule
\end{tabular}
\end{table}

@@

Figure~\ref{fig:planning_analysis} demonstrates DARIL's smooth degradation across horizons—from 33.6\% at 1s to 29.2\% at 10s (13.1\% relative decrease). This contrasts with RL approaches showing steeper degradation patterns, particularly world model RL exhibiting catastrophic performance loss.

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{planning_analysis.eps}
\caption{Planning performance comparison across methods and time horizons. DARIL maintains superior and stable performance while RL approaches show varying degrees of degradation. Error bars indicate 95\% confidence intervals.}
\label{fig:planning_analysis}
\end{figure}

@@

\textbf{Resource Allocation}: Research resources might be better allocated to optimizing IL architectures rather than developing complex RL systems for surgical planning tasks with expert demonstrations.

@@

Several limitations should be considered: (1) Single dataset evaluation on CholecT50 may not generalize to other surgical procedures. (2) Expert test data similar to training distributions may favor IL—results might differ with sub-expert or out-of-distribution scenarios. (3) Evaluation metrics directly reward expert-like behavior—alternative criteria focusing on patient outcomes might favor RL. (4) More sophisticated RL implementations with better state representations and reward design might outperform IL.

@@

%
\bibliographystyle{splncs04}
\bibliography{references}
